---
title: "Cyclistic bike-share analysis"
author: "Yehao Zheng"
date: "`r Sys.Date()`"
output: html_document
---

## About the conpany
Cyclistic is a fictional bike-share company launched in 2016. The shareable bikes are geotracked and locked into a network of 692 stations across Chicago. The bikes can be unlocked from one station and returned to any other station in the system anytime.
Cyclistic offers flexible pricing plans: single-ride passes, full-day passes, and annual memberships. Customers who purchase single-ride or full-day passes are referred to as casual riders. Customers who purchase annual memberships are Cyclistic members.

## Introduction
As a junior data analyst working in the marketing team of a bike-share company Cyclistic in Chicago, I'm responsible for collecting, analyzing, and reporting data that helps guide Cyclistic marketing strategy. 
In order to get the Cyclistic executives approval for my recommendations, I will come up with compelling data insights and professional data visualizations.
The project follow the data analysis process: ask, prepare, process, analyze, share, and act.

## Ask
To improve the company's performance, the director of marketing believes the company needs to maximize the number of annual merberships. Therefore, I want to understand How do annual members and casual riders use Cyclistic bikes differently, and designs a new marketing strategy to convert casual riders into annual members.

## Prepare
I will be using the [Cyclistic’s historical trip data](https://divvy-tripdata.s3.amazonaws.com/index.html) from May 2021 to April 2022 to analyze and identify trends. The data has been made available by Motivate International Inc. under this [license](https://ride.divvybikes.com/data-license-agreement). It is public data that you can use to explore how different customer types are
using Cyclistic bikes. 

## Process

### Load necessary libraries

```{r, echo=TRUE, message=FALSE, results = 'hide'}
library(tidyverse) # needed for read_csv()
library(janitor) # needed for compare_df_cols()
library(dplyr) # needed for bind_rows()
library(hms) # change difftime to HHMMSS
library(scales) # scale_y_continuous(labels = comma)
```

### Load the previous 12 months of cyclistic trip data 

```{r, echo=TRUE, message=FALSE, results = 'hide'}
data_202105 <- read_csv("/Users/yehao/Desktop/Coursera/Google Data Analytics Professional Certification/Case Study 1/data/202105-divvy-tripdata.csv")
data_202106 <- read_csv("/Users/yehao/Desktop/Coursera/Google Data Analytics Professional Certification/Case Study 1/data/202106-divvy-tripdata.csv")
data_202107 <- read_csv("/Users/yehao/Desktop/Coursera/Google Data Analytics Professional Certification/Case Study 1/data/202107-divvy-tripdata.csv")
data_202108 <- read_csv("/Users/yehao/Desktop/Coursera/Google Data Analytics Professional Certification/Case Study 1/data/202108-divvy-tripdata.csv")
data_202109 <- read_csv("/Users/yehao/Desktop/Coursera/Google Data Analytics Professional Certification/Case Study 1/data/202109-divvy-tripdata.csv")
data_202110 <- read_csv("/Users/yehao/Desktop/Coursera/Google Data Analytics Professional Certification/Case Study 1/data/202110-divvy-tripdata.csv")
data_202111 <- read_csv("/Users/yehao/Desktop/Coursera/Google Data Analytics Professional Certification/Case Study 1/data/202111-divvy-tripdata.csv")
data_202112 <- read_csv("/Users/yehao/Desktop/Coursera/Google Data Analytics Professional Certification/Case Study 1/data/202112-divvy-tripdata.csv")
data_202201 <- read_csv("/Users/yehao/Desktop/Coursera/Google Data Analytics Professional Certification/Case Study 1/data/202201-divvy-tripdata.csv")
data_202202 <- read_csv("/Users/yehao/Desktop/Coursera/Google Data Analytics Professional Certification/Case Study 1/data/202202-divvy-tripdata.csv")
data_202203 <- read_csv("/Users/yehao/Desktop/Coursera/Google Data Analytics Professional Certification/Case Study 1/data/202203-divvy-tripdata.csv")
data_202204 <- read_csv("/Users/yehao/Desktop/Coursera/Google Data Analytics Professional Certification/Case Study 1/data/202204-divvy-tripdata.csv")
```

### Examine the data types for the columns of each dataset, Ensure merge can be successful

```{r pressure}
compare_df_cols(data_202105,data_202106,data_202107,data_202108,data_202109,data_202110,data_202111,data_202112,data_202201,data_202202,data_202203,data_202204)
```

### Merge the previous 12 months of data

```{r, echo = TRUE, results = "hide"}
data_12months <- bind_rows(data_202105,data_202106,data_202107,data_202108,data_202109,data_202110,data_202111,data_202112,data_202201,data_202202,data_202203,data_202204)
```

### See if the merge is successful

```{r}
str(data_12months)
```

### Calculate the length of each ride in secs

```{r, echo = TRUE, results = "hide"}
ride_length <- difftime(data_12months$ended_at, data_12months$started_at, units = "secs")
```

### Change the length of each ride to the format of HHMMSS and store it in a new column ride_length

```{r, echo = TRUE, results = "hide"}
# x <- abs(as.numeric(ride_length))
# data_12months$ride_length <- sprintf("%02d:%02d:%02d",  x %% 86400 %/% 3600, x %% 3600 %/% 60,  x %% 60 %/% 1)
# data_12months$day_of_week <- weekdays(data_12months$started_at)

data_12months$ride_length <- as_hms(ride_length)
```

### Filter out the rows with ride_length <= 0

```{r}
data_12months <- filter(data_12months, ride_length > 0)
str(data_12months)
```

### Create a column “day_of_week" and calculate the day of the week that each ride started and select the necessary columns for analysis

```{r}
data_12months$day_of_week <- weekdays(data_12months$started_at)
data_12months <- data_12months %>% 
  select(ride_id, rideable_type, started_at, ended_at, member_casual, ride_length, day_of_week)
str(data_12months)
```

## Analyze

### Calculate the mean of ride_length

```{r}
length_secs <- as.numeric(data_12months$ride_length)
cat('Mean of ride_length is', mean(length_secs), "\n")
cat('Max ride_length is', max(length_secs), "\n")
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}
cat('Mode of day_of_week is', getmode(data_12months$day_of_week))
```

### Calculate the average ride_length for members and casual riders.

```{r}
mean_rl_mc <- data_12months %>% 
  group_by(member_casual) %>% 
  summarise("Average of ride_length" = round(mean(ride_length), 2)) %>% 
  rename("member or casual"= member_casual)
mean_rl_mc
```

### Sort the data from Monday to Sunday and then calculate the average ride_length for users by day_of_week.

```{r message=FALSE, warning=FALSE}
data_12months$day_of_week <- ordered(data_12months$day_of_week, levels=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
mean_rl_mc_wday <- data_12months %>% 
  group_by(day_of_week, member_casual) %>% 
  summarise("Average_of_ride_length" = round(mean(ride_length), 2)) 
mean_rl_mc_wday

```

```{r}
ggplot(data = mean_rl_mc_wday) + 
  geom_bar(mapping = aes(x = day_of_week, y = as.numeric(Average_of_ride_length), fill = member_casual),  stat='identity') +
  labs(x = "day of week", y = "average of ride length", title = "Average Ride Length vs Day of Week") +
  theme(plot.title = element_text(hjust = 0.5)) 
```

### Calculate the number of rides for users by day_of_week by adding Count of trip_id to Values.

```{r message=FALSE, warning=FALSE}
num_ride_wday <- data_12months %>% 
  group_by(day_of_week, member_casual) %>% 
  summarise("number_of_rides" = n_distinct(ride_id)) 
num_ride_wday
```

```{r}
ggplot(data = num_ride_wday) +
  geom_bar(mapping = aes(x = day_of_week, y = number_of_rides, fill = member_casual), stat =  "identity") +
  labs(x = "day of week", y = "number of rides", title = "Day of Week vs Number of Rides") +
  theme(plot.title = element_text(hjust = 0.5)) 
```

### Difference in number of rides for both annual member and casual riders by month

```{r message=FALSE, warning=FALSE}
#num_ride_per_day <- data_12months %>% 
#  group_by("day" = as.Date(started_at)) %>% 
#  summarise("number_of_rides" = n_distinct(ride_id))

num_ride_per_month <- data_12months %>% 
  group_by("month" = format(as.Date(started_at), "%Y-%m"), member_casual) %>% 
  summarise("number_of_rides" = n_distinct(ride_id))

ggplot(data = num_ride_per_month) + 
  geom_line(mapping = aes(x = month, y = number_of_rides, colour = member_casual, group = member_casual)) +
  scale_y_continuous(labels = comma) +
  labs(title = "Monthly Number of Rides") +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

### Difference in length of rides for both annual member and casual riders by month

```{r message=FALSE, warning=FALSE}
ride_length_per_month <- data_12months %>% 
  group_by("month" = format(as.Date(started_at), "%Y-%m"), member_casual) %>% 
  summarise("average_length_of_rides" = round(mean(ride_length), 2))
ride_length_per_month
ggplot(data = ride_length_per_month) + 
  geom_line(mapping = aes(x = month, y = average_length_of_rides, colour = member_casual, group = member_casual)) +
  scale_y_continuous(labels = comma) +
  labs(title = "Monthly Length of Rides") +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## Share

## Act

From the above tables and graphs, we know that the annual members tend to ride the bikes for commute to work because their usage of bikes are higher during the weekdays. In contrast, the casual riders prefer to ride the bikes for leisure during the weekend. we can also conclude that most of the time annual members have used the shareable bikes more frequently than the casual riders. However,the average ride length of casual riders are significantly longer than the annual members. 
To improve the company's performance, we can use this conclusion to convince the casual riders that even though they might not use our bikes very often, they would probably save more money by becoming a annual member because each of their ride is expensive based on the length. Then, we can also introduce a new annual weekend plan where the members of this plan can use the bikes freely during all the weekends.  





